Facility and Time Step Discretization Effects in Fuel Cycle Simulation

### 2015-09-29 paper.tex
more edits

All results data, analysis tools and instructions are publicly available for
download at \url{http://dx.doi.org/10.6084/m9.figshare.1546775}
\cite{Carlsen2015}.

@@

Figure \ref{fig:power} shows an overall view of the generated
power over time for each of the four cases.  Figure
\ref{fig:power-rel} normalizes the Figure
\ref{fig:power} curves to the expected exponential power curve
magnifying some interesting differences between the four cases.  For the first
100 years, the four cases behave somewhat similarly, although the individually
modeled reactors in cases 1 and 3 actually go online and offline for refueling
causing more variance.  Around year 100, a fuel shortage begins and more
significant differences between the four cases become apparent. This fuel
shortage lasts until about year 140. The power generated in Case 4 is lower
than in case 2 during the fuel shortage years. This is primarily due to the
\emph{quantized shutdown effect}.  The longer 3-month time step in case 4
results in reactor capacity going offline longer than necessary.

@@

\begin{figure*}[!h]
    \centering
    \includegraphics[width=0.8\textwidth]{exp2/power.eps}
    \caption[Generated power]{
        Generated power for all four cases.  A fuel shortage occurs from about
        year 100 to about year 140.  In cases 1 and 3 with individual reactor
        modeling, refueling cycles become much more synchronized during and
        after the fuel shortage resuting in large power fluctuations (the
        \emph{cycle staggering effect}).
    }
    \label{fig:power}
\end{figure*}

\begin{figure*}[!h]
    \centering
    \includegraphics[width=0.8\textwidth]{exp2/power-rel.eps}
    \caption[Normalized power]{
        Generated power is normalized to the expected 1\% exponential growth
        curve for all four cases.  A fuel shortage occurs from about year 100
        to about year 140.  The \emph{cycle staggering effect} can be seen as
        divergence from 1.0 and is especially visible in cases 1 and 3 where
        the shortage increases cycle synchronization significantly. The
        \emph{quantized shutdown effect} can also be seen in the discrepency
        between quarterly and monthly time steps for fleet reactor modeling
        (cases 2 and 4) during the shortage; a longer time step causes more
        reactor outage than necessary. The consistent jumping up and down for
        the fleet reactor cases is due to the mis-alignment of reactors going
        offline (multiples of 12 months) and the build period (21 months).
    }
    \label{fig:power-rel}
\end{figure*}

@@

Figure \ref{fig:unfueled} shows the fuel shortage more explicitly
for each of the four cases. The individual reactor modeling cases result in
more cumulative outage than corresponding fleet-based cases. The quarterly
time step of cases 3 and 4 make the outage significantly more severe. Cases 1
and 3 exhibit approximately twice as much offline power as their corresponding
fleet based simulations in cases 2 and 4.  The monthly time step simulations
also have roughly twice as much offline power as their corresponding quarterly
cases.

@@

Quantifying imperfect fuel sharing is a bit difficult, but Figure
\ref{fig:badshare} gives one way to approximate the \emph{fuel
sharing effect}.  Fleet reactor modeling by design exhibits perfect fuel
sharing by design (zero inefficiency) and so is not included in this figure.
Every batch that is given to a reactor that ends up not being able to start it
cycle at the scheduled time (due to fuel shortage) adds one to this cumulative
for each time step the cycle start is delayed.

One useful over-estimate of the fuel sharing inefficiency is to assume every
one of these batches could have been given to a reactor enabling it to
operate.  Multiplying each of these cumulative batches by reactor capacity
(i.e. 450 MWe) provides a way to compare fuel sharing inefficiency with the
overall fuel outage.  Correcting for capacity factor in is appropriate these
cases because the outage is many reactor cycle lengths in duration. This
results in a cumulative offline power caused by poor fuel sharing of roughly
1,600 $\text{GWe} \cdot \text{months}$ and 5,400 $\text{GWe} \cdot
\text{months}$ for cases 1 and 3 respectively.  These approximations are a
significant portion of the difference between individual and fleet cases in
Figure \ref{fig:unfueled} suggesting that most of the discrepency
in cumulative outage between fleet and individual modeling is caused by
inefficient fuel sharing in the individual reactor cases.

The poor fuel sharing exhibited in the individual reactor cases would not
occur if there were no new reactors being built during the shortage.
Previously operating reactors only ever need one batch - receiving only a
single batch allows them to operate.  Newly constructed reactors need multiple
batches in order to begin operation, and if they receive less than a full core
they cannot operate.  The real world doesn't necessarily optimize for fuel
sharing efficiency very cleanly. There are other factors that can affect fuel
sharing outcomes.  A few include:

@@

Having multiple assemblies per batch (item 1 above) will further degrade fuel
sharing efficiency.  Even reactors that are short only one batch could
potentially receive some fuel and still be unable to operate.  In order to
maintain optimal sharing, fresh fuel now needs to be sent in all-or-nothing
multi-assembly quanta with preference for reactors that need fewer assemblies.
Spare fresh fuel inventory (item 2 above) will have the global effect
increasing total outage severity; on average more batches of fresh fuel will
be sitting unused.  However, it can potentially reduce the frequency of
outages for individual reactors in some cases.

>>>>

All results data, analysis tools and instructions are publicly available for
download and use at \url{http://dx.doi.org/10.6084/m9.figshare.1546775}
\cite{Carlsen2015}.

@@

Figure \ref{fig:power} shows an overall view of the generated power over time
for each of the four cases.  Figure \ref{fig:power-rel} normalizes the Figure
\ref{fig:power} curves to the expected exponential power curve magnifying some
interesting differences between the four cases.  For the first 100 years, the
four cases behave somewhat similarly, although the individually modeled
reactors in cases 1 and 3 actually go offline and back online for refueling
causing more variance.  Around year 100, a fuel shortage begins and more
significant differences between the four cases become apparent. This fuel
shortage lasts until about year 140. The power generated in Case 4 is slightly
lower than in case 2 during the fuel shortage years. This is primarily due to
the \emph{quantized shutdown effect}.  The longer 3-month time step in case 4
results in reactor capacity going offline longer than necessary.

@@

\begin{figure*}[!h]
    \centering
    \includegraphics[width=0.8\textwidth]{exp2/power.eps}
    \caption[Generated power]{
        Generated power for all four cases.  A fuel shortage occurs from about
        year 100 to about year 140.  In cases 1 and 3 with individual reactor
        modeling, refueling cycles become much more synchronized during and
        after the fuel shortage resuting in large power fluctuations - an artifact of the
        \emph{cycle staggering effect}.
    }
    \label{fig:power}
\end{figure*}

\begin{figure*}[!h]
    \centering
    \includegraphics[width=0.8\textwidth]{exp2/power-rel.eps}
    \caption[Normalized power]{
        Generated power is normalized to the expected 1\% exponential growth
        curve for all four cases.  A fuel shortage occurs from about year 100
        to about year 140.  The \emph{cycle staggering effect} can be seen as
        excessive divergence above and below $1.0$ in cases 1 and 3 where
        the shortage increases cycle synchronization significantly. The
        \emph{quantized shutdown effect} can also be seen in the discrepency
        between quarterly and monthly time steps for fleet reactor modeling
        (cases 2 and 4) during the shortage; a longer time step causes more
        reactor outage than necessary. The consistent jumping up and down for
        the fleet reactor cases (even pre-shortage) is due to the mis-alignment of reactors going
        offline (multiples of 12 months) and the build period (21 months).
    }
    \label{fig:power-rel}
\end{figure*}

@@

Figure \ref{fig:unfueled} shows the fuel shortage more explicitly for each of
the four cases. The individual reactor modeling cases result in more
cumulative outage than corresponding fleet-based cases. The quarterly time
step of cases 3 and 4 also make the outage significantly more severe. Cases 1
and 3 exhibit approximately twice as much offline power as their corresponding
fleet based simulations in cases 2 and 4.  The monthly time step simulations
also have roughly twice as much offline power as their corresponding quarterly
cases.

@@

Quantifying imperfect fuel sharing is a bit difficult, but Figure
\ref{fig:badshare} gives one way to approximate the \emph{fuel
sharing effect}.  Fleet reactor modeling by design exhibits perfect fuel
sharing by design (zero inefficiency) and so is not included in this figure.
Every batch that is given to a reactor that ends up not being able to start its
cycle at the scheduled time (due to fuel shortage) adds one to this cumulative
for each time step the cycle start is delayed.

One useful over-estimate of the fuel sharing inefficiency is to assume every
one of the "wasted" batches could have been given to a reactor enabling it to
operate.  Multiplying each of these cumulative batches by reactor capacity
(i.e. 450 MWe) provides a way to compare fuel sharing inefficiency with the
overall fuel outage.  This results in a cumulative offline power caused by
poor fuel sharing of roughly 2,000 $\text{GWe} \cdot \text{months}$ and 6,600
$\text{GWe} \cdot \text{months}$ for cases 1 and 3 respectively.  These
approximations are a significant portion of the difference between individual
and fleet cases in Figure \ref{fig:unfueled} suggesting that most of the
discrepency in cumulative outage between fleet and individual modeling is
caused by inefficient fuel sharing in the individual reactor cases.

The poor fuel sharing exhibited in the individual reactor cases would not
occur if there were no new reactors being built during the shortage.
Previously operating reactors only ever need one batch - receiving only a
single batch allows them to operate.  Newly constructed reactors need multiple
batches in order to begin operation, and if they receive less than a full core
they cannot operate.  The real world doesn't necessarily optimize for fuel
sharing efficiency very cleanly. There are many factors that can affect
real-world fuel sharing outcomes.  A few include:

@@

Having multiple assemblies per batch (item 1 above) will further degrade fuel
sharing efficiency.  Even reactors that are short only one batch could
potentially receive some fuel and still be unable to operate.  In order to
maintain optimal sharing, fresh fuel now needs to be sent in all-or-nothing
multi-assembly quanta with higher preference to reactors that need fewer
assemblies.  Spare fresh fuel inventory (item 2 above) will have the global
effect of increasing total outage severity; on average more batches of fresh
fuel will be sitting unused.  However, it can potentially reduce the frequency
of outages for individual reactors in some cases.
